Forecasting the age structure of the scientific workforce in Australia

### 2025-08-22 age_structure_forecasts.qmd
Working on paper

In planning for the future labour market, it is necessary to forecast the age structure of the workforce, in order to enable informed decision-making on policies, especially concerning universities and immigration. We propose a statistical modelling approach to this problem, illustrated using various scientific disciplines in Australia, forecasting future workforce age structures over the next decade. The forecasts described have been used by the Australian Academy of Science as part of _Australian Science, Australia’s Future: Science 2035_, an initiative assessing the capability of the national science system and its role in achieving Australia’s ambitions [@AAS2035].

@@

We describe the methodology in @sec-methodology. By way of illustration, we apply the methodology to major scientific disciplines in Australia, focusing on the Natural and Physical Sciences. We describe the data sources in @sec-data, with the results provided in @sec-results. The aim of this analysis is to inform future workforce planning and policy decisions to support the growth of Australia’s scientific community. Finally, we provide some discussion and conclusions in @sec-conclusion.

@@

Suppose our workforce is divided into $I$ groups, indexed by $i=1,\dots,I$. In our application, these are scientific disciplines, but in principle they could refer to any subdivision of workers. Let $P_{i,x,t}$ denote the number of equivalent full-time workers in group $i$ who are aged $x$ at the start of year $t$, where $x=15,16,\dots$. The starting age of 15 is chosen because it is the minimum age at which individuals are counted as part of the labour force in the Australian Census [@ABSLFSP]. Forecasts are required for $P_{i,x,T+h}$ across all ages and groups, for some forecast horizon $h>0$.

People can leave the workforce of a group through death, retirement, career change, or emigration. They can enter the workforce through graduation, immigration, or career change. Unfortunately, we typically do not have data on many of these processes, so we will combine career change and migration into a remainder term, which we denote as $E_{i,x,t}$. We usually do have data on numbers of deaths ($D_{i,x,t}$), numbers of retirements ($R_{i,x,t}$), and numbers of graduates ($G_{i,x,t}$); the numbers in each case are for people aged $x$ at the start of year $t$.

The population can be forecasted using a model similar to the stochastic population model of @HB08, excluding the birth process:
$$
P_{i,x+1,t+1} = P_{i,x,t} - D_{i,x,t} - R_{i,x,t} + G_{i,x,t} + E_{i,x,t}, \quad x = 15,16,\dots
$$ {#eq-dgb}
That is, the population each year is equal to the population from the previous year having aged 1 year, minus the deaths or retirements that occurred during the previous year, plus the new graduates, plus any other changes due to migration or career change (which may be negative). We assume that $D$, $R$ and $G$ are all non-negative, $D_{i,x,t} + R_{i,x,t} \le P_{i,x,t}$, and $E_{i,x,t} = G_{i,x,t} = 0$ above some age threshold (say $x=100$). Once $P_{i,x,t} = 0$ when $x$ is above that threshold, all future populations $P_{i,x+k,t+k} = 0$, $k=1,2,\dots$. That is, when the cohort aged $x$ in year $t$ has all retired or died, they will not be replaced through new graduates or migrants of the same age.

@@

To use @eq-dgb, data on $P$, $D$, $R$, and $G$ is required, and the remainder component $E$ will be calculated using @eq-dgb. It is unlikely that we have available separate death and retirement counts for each group, so we will instead use probabilities of death and retirement across all groups, giving $D_{i,x,t} = P_{i,x,t}q_{x,t}$ and $R_{i,x,t} = P_{i,x,t}r_{x,t}$ where $q_{x,t}$ is the probability of death at each $x$ in year $t$, and $r_{x,t}$ is the probability of retirement at each $x$ in year $t$. Similarly, graduation numbers are rarely available by discipline and age, so we will approximate $G_{i,x,t} = g_xG_t$ where $G_{i,t}$ is the total number of graduates in year $t$ and $g_x$ is the proportion of graduates by age across all disciplines. So our more realistic model becomes
$$
P_{i,x+1,t+1} = P_{i,x,t}(1 - q_{x,t} - r_{x,t}) + g_xG_{i,t} + E_{i,x,t}, \quad x = 15,16,\dots
$$ {#eq-dgb2}
To obtain forecasts of $P$, forecasts of $q$, $r$, $G$, and $E$ must then be computed. Each of these can be estimated using a functional data model [@HU07].

@@

Data on $P_{i,x,t}$ are sourced from the *Census of Population and Housing* (2006, 2011, 2016, and 2021) [@ABSCensus]. This dataset encompasses labour force participation status, age group, the highest level of completed non-school qualification level (QALLP), the corresponding field of study [QALFP; @ABSQALFP], and the industries in which individuals work.

However, labour force participation data is available only for 2016 and 2021 [@ABSLFFP]. To estimate worker numbers for 2006 and 2011, the average participation rates from 2016 and 2021 were applied, assuming overall patterns remain consistent. While participation rates were generally stable between 2016 and 2021, smaller groups experienced more volatility. Younger and older age groups within some disciplines were particularly affected by this volatility, especially when sample sizes were small. For example, in a group of five 15--19-year-olds, a single individual's employment status change could significantly shift the participation rate. Similarly, fluctuations in smaller cohorts of older scientists impacted participation rates. Averaging the rates from 2016 and 2021 helps mitigate this volatility and provides a more stable estimate for 2006 and 2011. Any discrepancies in participation rates are minimal due to the small cohort sizes, making this a more reliable approach than relying on a single year.

The census data on scientists active in the labour market, provided by the ABS in one-year age groups, is shown in @fig-cohort-interpolation as the thick lines. As data is only available in census years (2006, 2011, 2016, 2021), cohort interpolation is used to estimate values for the intervening years. Cohort interpolation methods [@STUPP88] apply linear interpolation within each age cohort between census years.

@@

Retirement intentions data is categorised by the industry of an individual's main job and is provided in broad age groups, as shown in @fig-beforedisaggregation. While the ABS dataset covers 19 industries, only the key industries relevant to this study are displayed in the figure.

To estimate single-year age distributions, a monotonic cubic spline [@SHW04] is applied directly to the cumulative values of these age groups, ensuring a smooth transition between them. This method preserves the overall distribution while generating continuous single-year estimates, as shown in @fig-retirements.

@@

Graduate completion statistics from 2006 to 2023, categorised by age group and including individuals with a bachelor’s degree or higher, are derived from the *Award Course Completions* dataset [@DeptEdu]. Additionally, the Department of Education provides data on the number of graduates with a bachelor’s degree or higher, categorised by discipline and year, for the same period [@DeptEdu_Private]. These datasets include both domestic and international students.

@@

This analysis contributes to the Australian Academy of Science’s initiative _Australian Science, Australia’s Future: Science 2035_ by identifying shifts in the workforce, helping anticipate future demand for science capabilities, and highlighting potential workforce gaps as aging and retirement reshape workforce dynamics.

>>>>

In planning for the future labour market, it is necessary to forecast the age structure of the workforce, in order to enable informed decision-making on policies, especially concerning universities and immigration. We propose a statistical modelling approach to this problem, illustrated using various scientific disciplines in Australia, forecasting future workforce age structures over the next decade. The forecasts described have been used by the Australian Academy of Science as part of _Australian Science, Australia's Future: Science 2035_, an initiative assessing the capability of the national science system and its role in achieving Australia's ambitions [@AAS2035].

@@

We describe the methodology in @sec-methodology. By way of illustration, we apply the methodology to major scientific disciplines in Australia, focusing on the Natural and Physical Sciences. We describe the data sources in @sec-data, with the results provided in @sec-results. The aim of this analysis is to inform future workforce planning and policy decisions to support the growth of Australia's scientific community. Finally, we provide some discussion and conclusions in @sec-conclusion.

@@

Suppose our workforce is divided into $I$ groups, indexed by $i=1,\dots,I$. In our application, these are scientific disciplines, but in principle they could refer to any subdivision of workers. Let $P_{i,x,t}$ denote the number of equivalent full-time workers in group $i$ who are aged $x$ at the start of year $t$, where $x=15,16,\dots$. The starting age of 15 is chosen because it is the minimum age at which individuals are counted as part of the labour force in the Australian Census [@ABSLFSP]. We assume that data are available for years $t=1,\dots,T$, and that forecasts are required for $P_{i,x,T+h}$ across all ages and groups, for some forecast horizon $h>0$.

People can leave the workforce of a group through death, retirement, career change, or emigration. They can enter the workforce through graduation, immigration, or career change. Unfortunately, we typically do not have data on many of these processes, so we will combine career change, emigration and immigration into a remainder term, which we denote as $E_{i,x,t}$. Suppose we do have data on numbers of deaths ($D_{i,x,t}$), numbers of retirements ($R_{i,x,t}$), and numbers of graduates ($G_{i,x,t}$); the numbers in each case are for people aged $x$ at the start of year $t$. Then population changes can be described using a model similar to the stochastic population model of @HB08, excluding the birth process:
$$
P_{i,x+1,t+1} = P_{i,x,t} - D_{i,x,t} - R_{i,x,t} + G_{i,x,t} + E_{i,x,t}, 
$$ {#eq-dgb}
That is, the population each year is equal to the population from the previous year having aged 1 year, minus the deaths or retirements that occurred during the previous year, plus the new graduates, plus any other changes due to migration or career change (which may be negative). We assume that $D$, $R$ and $G$ are all non-negative, $D_{i,x,t} + R_{i,x,t} \le P_{i,x,t}$, and $E_{i,x,t} = G_{i,x,t} = 0$ above some age threshold (say $x=100$). Once $P_{i,x,t} = 0$ when $x$ is above that threshold, all future populations $P_{i,x+k,t+k} = 0$, forw $k=1,2,\dots$. That is, when the cohort aged $x$ in year $t$ has all retired or died, they will not be replaced by new workers of the same age.

@@

To use @eq-dgb, data on $P$, $D$, $R$, and $G$ is required, and the remainder component $E$ will be calculated using @eq-dgb. It is unlikely that we have available separate death and retirement counts for each group, so we will instead use probabilities of death and retirement across all groups, giving $D_{i,x,t} = P_{i,x,t}q_{x,t}$ and $R_{i,x,t} = P_{i,x,t}r_{x,t}$ where $q_{x,t}$ is the probability of death at each $x$ in year $t$, and $r_{x,t}$ is the probability of retirement at each $x$ in year $t$. Similarly, graduation numbers are rarely available by discipline and age, so we will approximate $G_{i,x,t} = g_xG_t$ where $G_{i,t}$ is the total number of graduates in year $t$ and $g_x$ is the proportion of graduates by age across all disciplines. So our more realistic model becomes
$$
P_{i,x+1,t+1} = P_{i,x,t}(1 - q_{x,t} - r_{x,t}) + g_xG_{i,t} + E_{i,x,t}, 
$$ {#eq-dgb2}
To obtain forecasts of $P$, forecasts of $q$, $r$, $G$, and $E$ must then be computed. Each of these can be estimated using a functional time series model [@HU07].

@@

Data on $P_{i,x,t}$ are sourced from the *Census of Population and Housing* (2006, 2011, 2016, and 2021) [@ABSCensus]. This dataset encompasses labour force participation status, age group, the highest level of completed non-school qualification level (QALLP), the corresponding field of study [QALFP, @ABSQALFP], and the industries in which individuals work. However, labour force participation data is available only for 2016 and 2021 [@ABSLFFP]. To estimate worker numbers for 2006 and 2011, the average participation rates from 2016 and 2021 were applied, assuming overall age distributions remain consistent.

The census data on scientists active in the labour market, provided by the ABS in one-year age groups, is shown in @fig-cohort-interpolation as the thick lines. As data is only available in census years (2006, 2011, 2016, 2021), cohort interpolation is used to estimate values for the intervening years (shown as thin lines). Cohort finterpolation methods [@STUPP88] apply linear interpolation within each age cohort between census years.

@@

Retirement intentions data are categorised by the industry of an individual's main job and is provided in four broad age groups (45--59, 60--64, 65--69 and 70+). While ABS data covers 19 industries, only the key industries relevant to science are included, namely Professional, scientific and technical services, Education and training, and Health care and social assistance.

To estimate a single-year retirement age distribution, a monotonic cubic spline [@SHW04] is applied directly to the cumulative values of these age groups, ensuring a smooth transition between them. This method preserves the overall distribution while generating continuous single-year estimates, as shown in @fig-retirements.

@@

Graduate completion statistics from 2006 to 2023, categorised by age group and including individuals with a bachelor's degree or higher, are derived from the *Award Course Completions* dataset [@DeptEdu]. Additionally, the Department of Education provides data on the number of graduates with a bachelor's degree or higher, categorised by discipline and year, for the same period [@DeptEdu_Private]. These datasets include both domestic and international students.

@@

This analysis contributes to the Australian Academy of Science's initiative _Australian Science, Australia's Future: Science 2035_ by identifying shifts in the workforce, helping anticipate future demand for science capabilities, and highlighting potential workforce gaps as aging and retirement reshape workforce dynamics.
